---
title: "Distance-based $R^2$: a controlled comparison of joint and sequential calculations"
author: "Bashir Hamidi"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    encoding: UTF-8
    code_folding: show
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    fig_width: 9
    fig_height: 6
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 3
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
  error = FALSE, cache = FALSE, dpi = 120)
options(width = 100)
set.seed(843)
```

```{css, echo=FALSE}
#TOC { max-width: fit-content; white-space: nowrap; }
div:has(> #TOC) { display: flex; flex-direction: row-reverse; }
.callout { border-left: 4px solid #2c7fb8; background: #f3f8fb;
  padding: 0.7em 1em; margin: 1em 0; }
.warning { border-left-color: #d95f0e; background: #fff7ec; }
```

# Purpose and main conclusion

This document uses one controlled simulated microbiome dataset as its running example.
The observations and predictors stay fixed while the distance measure changes. This
separates two issues that were difficult to distinguish in the original Enterotype
example:

1. **Sequential-projection issue:** fitting $Z$, reconstructing a residual distance,
   and then fitting the original $F$ is not the standard nested comparison
   $Z\subset Z+F$.
2. **Negative-eigenvalue-removal issue:** `a.dist()` replaces negative eigenvalues of
   each model-specific residual matrix with zero before reconstructing a Euclidean
   distance. The resulting residual sums of squares need not partition the original
   distance geometry.

The first issue is general and remains even when the starting distance is Euclidean.
The second depends on the distance measure. It is prominent for non-Euclidean inputs
and disappears, apart from numerical rounding, for Euclidean inputs.

The Enterotype data and the variables `SeqTech`, `Enterotype`, and `Age` are retained
later as real-data illustrations rather than used to organize the document.

<div class="callout warning">
The recommended calculation is to obtain $TSS$, $RSS_Z$, and $RSS_{ZF}$ from one
shared Gower matrix. The additional contribution of $F$ after $Z$ is the nested
contrast $RSS_Z-RSS_{ZF}$. It is not obtained by calling `a.dist()` twice.
</div>

# Computational setup

```{r installation, eval=FALSE}
install.packages("remotes")
remotes::install_github("bashirhamidi/WdStar")
```

```{r packages}
library(WdStar)
library(phyloseq)
```

The helper below inspects a centered Gower matrix before `a.dist()` removes negative
eigenvalues. The negative absolute fraction is

$$
\frac{\sum_{\lambda_i<0}|\lambda_i|}{\sum_i|\lambda_i|}.
$$

```{r helpers}
as_number <- function(x) unname(as.numeric(x))

residual_eigen_summary <- function(dm, formula = NULL, data = NULL, tol = 1e-8) {
  d2 <- as.matrix(dm)^2
  n <- nrow(d2)
  J <- diag(n) - matrix(1 / n, nrow = n, ncol = n)
  G <- -0.5 * J %*% d2 %*% J
  if (!is.null(formula)) {
    mf <- model.frame(formula, data = data, drop.unused.levels = TRUE)
    X <- model.matrix(formula, data = mf)
    qx <- qr(X)
    Q <- qr.Q(qx)[, seq_len(qx$rank), drop = FALSE]
    M <- diag(n) - tcrossprod(Q)
    G <- M %*% G %*% M
  }
  values <- eigen((G + t(G)) / 2, symmetric = TRUE, only.values = TRUE)$values
  threshold <- tol * max(1, max(abs(values)))
  positive <- values > threshold
  negative <- values < -threshold
  positive_sum <- sum(values[positive])
  negative_sum <- sum(values[negative])
  data.frame(
    positive_n = sum(positive), negative_n = sum(negative),
    near_zero_n = sum(!(positive | negative)), positive_sum = positive_sum,
    negative_sum = negative_sum, signed_sum = positive_sum + negative_sum,
    positive_only_sum = positive_sum,
    negative_absolute_fraction =
      abs(negative_sum) / (positive_sum + abs(negative_sum))
  )
}
```

# Definitions

Let $TSS$ be the total distance-based sum of squares, $RSS_F$ the residual sum of
squares from $F$ alone, $RSS_Z$ the residual sum of squares from $Z$ alone, and
$RSS_{ZF}$ the residual sum of squares from the full model containing $Z$ and $F$.

## Standalone, adjustment, and full-model fit

$$
R_F^2=1-\frac{RSS_F}{TSS},\qquad
R_Z^2=1-\frac{RSS_Z}{TSS},\qquad
R_{Z+F}^2=1-\frac{RSS_{ZF}}{TSS}.
$$

$R_F^2$ is the standalone fit of $F$. $R_Z^2$ is the adjustment-only fit.
$R_{Z+F}^2$ is the collective fit of the full model and does not isolate $F$.

## Additional contribution of $F$ after $Z$

$$
SS_{F\mid Z}=RSS_Z-RSS_{ZF}.
$$

The semi-partial contribution uses the original total variation:

$$
R_{\mathrm{semi},F\mid Z}^2
=\frac{RSS_Z-RSS_{ZF}}{TSS}=R_{Z+F}^2-R_Z^2.
$$

It answers: after accounting for $Z$, what additional proportion of the *original*
variation is explained by adding $F$?

The partial contribution uses the variation remaining after $Z$:

$$
R_{\mathrm{partial},F\mid Z}^2
=\frac{RSS_Z-RSS_{ZF}}{RSS_Z}
=\frac{R_{Z+F}^2-R_Z^2}{1-R_Z^2}.
$$

It answers: of the variation *remaining after $Z$*, what proportion is explained by
adding $F$? The cumulative fit is
$R_Z^2+R_{\mathrm{semi},F\mid Z}^2=R_{Z+F}^2$.

When $Z$ and $F$ overlap, sequential term contributions depend on entry order. A
correctly computed final model with the same columns does not: joint and proper
sequential fitting have the same $RSS_{ZF}$.

# Primary running example: controlled simulation

## Design

The simulation has 120 samples and 100 taxa. Both $Z$ and $F$ are balanced binary
variables, but they agree for 104 of the 120 samples. They are strongly, but not
perfectly, associated. Taxon effects include structure shared by $Z$ and $F$,
opposing effects that produce suppression, and additional $F$-specific structure.
Counts are drawn at a depth of 5,000 reads per sample.

This is a stress test chosen to make the computational differences visible. It is not
intended to represent a typical microbiome effect size.

```{r simulation}
simulate_microbiome <- function(seed = 1103, n = 120L, taxa = 100L,
  discordant_per_z_level = 8L, sequencing_depth = 5000L) {
  set.seed(seed)
  z01 <- rep(c(0, 1), each = n / 2L)
  f01 <- z01
  for (z_value in 0:1) {
    candidates <- which(z01 == z_value)
    flip <- sample(candidates, discordant_per_z_level)
    f01[flip] <- 1L - f01[flip]
  }
  z_centered <- z01 - mean(z01)
  f_centered <- f01 - mean(f01)
  baseline <- rnorm(taxa, mean = -2.5, sd = 0.8)
  beta_z <- beta_f <- numeric(taxa)
  beta_z[1:15] <- 1; beta_z[16:30] <- -1
  beta_f[1:15] <- -2; beta_f[16:30] <- 2
  beta_f[31:45] <- 1; beta_f[46:60] <- -1

  counts <- matrix(0L, nrow = taxa, ncol = n)
  for (i in seq_len(n)) {
    log_abundance <- baseline + z_centered[i] * beta_z +
      f_centered[i] * beta_f + rnorm(taxa, sd = 0.4)
    probabilities <- exp(log_abundance - max(log_abundance))
    probabilities <- probabilities / sum(probabilities)
    counts[, i] <- as.vector(rmultinom(1L, sequencing_depth, probabilities))
  }
  rownames(counts) <- paste0("Taxon", seq_len(taxa))
  colnames(counts) <- paste0("Sample", seq_len(n))
  metadata <- data.frame(
    Z = factor(z01, labels = c("Z0", "Z1")),
    F = factor(f01, labels = c("F0", "F1")), row.names = colnames(counts)
  )
  phyloseq(otu_table(counts, taxa_are_rows = TRUE), sample_data(metadata))
}

simulated_data <- simulate_microbiome()
simulated_metadata <- data.frame(sample_data(simulated_data))
simulated_otu <- as(otu_table(simulated_data), "matrix")
if (taxa_are_rows(simulated_data)) simulated_otu <- t(simulated_otu)
```

## Why test the association between $Z$ and $F$?

```{r simulation-association}
simulated_fz_table <- table(Z = simulated_metadata$Z, F = simulated_metadata$F)
simulated_chisq <- chisq.test(simulated_fz_table, correct = FALSE)
simulated_cramers_v <- sqrt(as_number(simulated_chisq$statistic) /
  (sum(simulated_fz_table) *
     min(nrow(simulated_fz_table) - 1, ncol(simulated_fz_table) - 1)))

knitr::kable(simulated_fz_table, caption = "Known simulated relationship between Z and F")
knitr::kable(data.frame(
  chi_squared = as_number(simulated_chisq$statistic),
  degrees_of_freedom = as_number(simulated_chisq$parameter),
  p_value = format.pval(simulated_chisq$p.value, digits = 3, eps = 0.001),
  Cramers_V = simulated_cramers_v), digits = 4)
```

The chi-square test asks whether the cross-tabulation differs from independence. It
does not test the distance matrix or microbiome model. Cramér's $V$ describes the
strength of predictor association. Here $V=$
`r sprintf("%.3f", simulated_cramers_v)`, confirming that the predictor spaces
overlap while both variables remain estimable. This is the setting in which
sequential allocation is consequential.

# Why one pass and two passes are different

Let $G$ be the Gower matrix obtained from the starting distance, and let $M_Z$ and
$M_{ZF}$ be residual projectors. A joint full-model projection is

$$E_{ZF}=M_{ZF}GM_{ZF}.$$

A proper sequential analysis compares $Z\subset Z+F$ in this same geometry. The
second step adds only the part of $F$ not represented by $Z$,
$F_{\perp Z}=M_ZF$. The final residual is still $E_{ZF}$.

The following two-pass code is different:

```r
D_Z <- a.dist(D, ~ Z, formula_data = metadata)
a.dist(D_Z, ~ F, formula_data = metadata)
```

It reconstructs a new distance after fitting $Z$ and then fits the original $F$ to
that changed response. Algebraically this resembles $M_FM_Z$, which is not generally
$M_{ZF}$ when $F$ and $Z$ overlap. The projections do not generally commute.

If $\Pi_+(A)$ means replacing negative eigenvalues with zero, the paths are

$$\text{one pass: }\Pi_+(M_{ZF}GM_{ZF})$$

and

$$\text{two passes: }\Pi_+\{M_F\Pi_+(M_ZGM_Z)M_F\}.$$

Once negative eigencomponents are removed after the first pass, the second pass
cannot recover them.

## What “removing negative eigenvalues” means

If $E=U\Lambda U^{\mathsf T}$, `a.dist()` replaces
$\lambda_i$ by $\lambda_i^+=\max(\lambda_i,0)$ and reconstructs a Euclidean
distance from $U\Lambda_+U^{\mathsf T}$. Earlier drafts called this *truncation*.
It does not remove samples, taxa, predictors, or pairwise distances. It removes the
negative axes of the residual geometry.

Before removal, residual SS is the signed trace

$$\operatorname{tr}(E)=\sum_{\lambda_i>0}\lambda_i+
\sum_{\lambda_i<0}\lambda_i.$$

After removal, it becomes $\sum_{\lambda_i>0}\lambda_i$, a larger quantity whenever
substantive negative eigenvalues were present. Model-specific reconstructed distances
therefore need not partition the original $TSS$.

# Bray--Curtis results for the simulation

```{r simulation-bray-models}
D_bray <- distance(simulated_data, method = "bray")
D_bray_Z <- a.dist(D_bray, ~ Z, simulated_metadata)
D_bray_full <- a.dist(D_bray, ~ Z + F, simulated_metadata)
D_bray_two <- a.dist(D_bray_Z, ~ F, simulated_metadata)
bray_adonis <- vegan::adonis2(D_bray ~ Z + F, data = simulated_metadata,
  by = "terms", permutations = 999)

bray_r2_z <- as_number(dist.goodness.of.fit(D_bray, D_bray_Z))
bray_r2_one <- as_number(dist.goodness.of.fit(D_bray, D_bray_full))
bray_r2_two <- as_number(dist.goodness.of.fit(D_bray, D_bray_two))
bray_vegan_z <- as_number(bray_adonis["Z", "R2"])
bray_vegan_semi <- as_number(bray_adonis["F", "R2"])
bray_vegan_residual <- as_number(bray_adonis["Residual", "R2"])
bray_vegan_full <- 1 - bray_vegan_residual

bray_comparison <- data.frame(
  quantity = c("Adjustment-only R2: Z", "Full/cumulative R2: Z + F",
    "Semi-partial R2: F after Z", "Partial R2: F after Z",
    "Residual proportion after full model"),
  vegan_shared_geometry = c(bray_vegan_z, bray_vegan_full,
    bray_vegan_semi, bray_vegan_semi / (1 - bray_vegan_z),
    bray_vegan_residual),
  WdStar_one_pass = c(bray_r2_z, bray_r2_one,
    bray_r2_one - bray_r2_z,
    (bray_r2_one - bray_r2_z) / (1 - bray_r2_z), 1 - bray_r2_one),
  WdStar_two_pass = c(bray_r2_z, bray_r2_two,
    bray_r2_two - bray_r2_z,
    (bray_r2_two - bray_r2_z) / (1 - bray_r2_z), 1 - bray_r2_two)
)
knitr::kable(bray_comparison, digits = 4)
```

The WdStar one-pass and two-pass full-model coefficients differ by
`r sprintf("%.4f", bray_r2_one - bray_r2_two)`. WdStar one-pass also differs from
vegan because the one-pass residual matrix is reconstructed after negative
eigenvalues are removed.

```{r simulation-bray-eigenvalues}
bray_eigen <- rbind(
  original = residual_eigen_summary(D_bray),
  z_before_removal = residual_eigen_summary(D_bray, ~ Z, simulated_metadata),
  full_before_removal = residual_eigen_summary(D_bray, ~ Z + F, simulated_metadata),
  second_pass = residual_eigen_summary(D_bray_Z, ~ F, simulated_metadata)
)
knitr::kable(data.frame(
  matrix = c("Original Bray-Curtis",
    "Z residual before negative eigenvalues are removed",
    "Joint Z + F residual before negative eigenvalues are removed",
    "F residual after reconstructed Z distance is supplied to pass 2"),
  positive_eigenvalues = bray_eigen$positive_n,
  negative_eigenvalues = bray_eigen$negative_n,
  signed_sum = bray_eigen$signed_sum,
  positive_only_sum = bray_eigen$positive_only_sum,
  negative_absolute_fraction = bray_eigen$negative_absolute_fraction), digits = 5)
```

No substantive negative eigenvalues appear in the second-pass row because the first
`a.dist()` call already removed them when it created `D_bray_Z`.

# Does the conclusion change with JSD or Jaccard?

The name of a method is not enough; the exact object passed to the model matters.

- Bray--Curtis and quantitative Jaccard (Ružička for abundance data) can be
  non-Euclidean.
- `vegan::vegdist(..., method = "jaccard", binary = TRUE)` computes incidence
  Jaccard. Jaccard is a metric, but its direct PCoA representation can contain
  negative eigenvalues. In this simulation, its square root is Euclidean.
- The installed `phyloseq` method `"jsd"` returns **Jensen--Shannon divergence**, not
  its square root. Square-root Jensen--Shannon is the metric, Hilbert-space form.
  Treating raw divergence as a distance causes the Gower calculation to square the
  divergence again.

```{r distance-sensitivity}
evaluate_distance <- function(label, dm, metadata) {
  D_Z <- a.dist(dm, ~ Z, metadata)
  D_full <- a.dist(dm, ~ Z + F, metadata)
  D_two <- a.dist(D_Z, ~ F, metadata)
  ad <- vegan::adonis2(dm ~ Z + F, data = metadata,
    by = "terms", permutations = 0)
  r2_one <- as_number(dist.goodness.of.fit(dm, D_full))
  r2_two <- as_number(dist.goodness.of.fit(dm, D_two))
  vegan_full <- 1 - as_number(ad["Residual", "R2"])
  data.frame(
    distance_definition = label,
    negative_absolute_fraction =
      residual_eigen_summary(dm)$negative_absolute_fraction,
    vegan_full_R2 = vegan_full,
    WdStar_one_pass_full_R2 = r2_one,
    WdStar_two_pass_full_R2 = r2_two,
    one_pass_minus_vegan = r2_one - vegan_full,
    one_pass_minus_two_pass = r2_one - r2_two)
}

D_quantitative_jaccard <- distance(simulated_data, method = "jaccard")
D_binary_jaccard <- vegan::vegdist(simulated_otu,
  method = "jaccard", binary = TRUE)
D_js_divergence <- distance(simulated_data, method = "jsd")
D_sqrt_js <- sqrt(D_js_divergence)
D_sqrt_binary_jaccard <- sqrt(D_binary_jaccard)

distance_sensitivity <- do.call(rbind, list(
  evaluate_distance("Bray-Curtis", D_bray, simulated_metadata),
  evaluate_distance("Quantitative Jaccard (Ruzicka)",
    D_quantitative_jaccard, simulated_metadata),
  evaluate_distance("Binary Jaccard", D_binary_jaccard, simulated_metadata),
  evaluate_distance("phyloseq jsd: raw Jensen-Shannon divergence",
    D_js_divergence, simulated_metadata),
  evaluate_distance("Square-root Jensen-Shannon divergence",
    D_sqrt_js, simulated_metadata),
  evaluate_distance("Square-root binary Jaccard",
    D_sqrt_binary_jaccard, simulated_metadata)
))
knitr::kable(distance_sensitivity, digits = 4)
```

The table separates the conclusions:

1. **The two-pass problem does not disappear with a Euclidean distance.** For
   square-root Jensen--Shannon, the negative fraction is zero and vegan agrees with
   WdStar one-pass, yet one-pass and two-pass differ by
   `r sprintf("%.4f", distance_sensitivity[5, "one_pass_minus_two_pass"])`.
2. **Vegan versus WdStar one-pass disagreement tracks non-Euclidean geometry.** It is
   absent for the two square-root distances here and present for Bray--Curtis,
   quantitative Jaccard, binary Jaccard, and raw JSD.
3. **Raw `phyloseq` JSD is unsuitable here without clarification.** It is the
   divergence itself. Its negative fraction is
   `r sprintf("%.3f", distance_sensitivity[4, "negative_absolute_fraction"])` and
   its reconstructed-distance coefficient differs greatly from vegan's signed
   pseudo-$R^2$. If Jensen--Shannon is desired, name and use the square-root form.
4. **“Jaccard” must be labeled quantitative or binary.** They answer different
   ecological questions. These simulated counts are dense, so binary Jaccard sees
   much less modeled structure than abundance-sensitive distances.

These values are properties of this simulation, not universal rankings. Changing the
distance changes the response geometry, $TSS$, and every model $R^2$. The general
rule is to inspect the exact definition and eigenvalues and calculate all nested sums
of squares in one shared geometry.

<div class="callout">
Switching from Bray--Curtis to Jaccard or Jensen--Shannon does not rescue the naive
two-pass algorithm. A Euclidean choice removes the negative-eigenvalue-removal part
of the discrepancy, but it does not make fitting the original $F$ to a previously
residualized distance equivalent to fitting $Z+F$.
</div>

# Lingoes-corrected Bray--Curtis as a diagnostic control

If $\lambda_{\min}$ is the smallest eigenvalue of the original Gower matrix, Lingoes
uses $c=\max(-\lambda_{\min},0)$ and

$$d_{ij}^{\mathrm{Lingoes}}=\sqrt{d_{ij}^2+2c},\qquad i\ne j,$$

with a zero diagonal. It changes all off-diagonal distances by one common rule and
defines a new Euclidean response geometry. It does not recover or improve the
original Bray--Curtis $R^2$. Here it is applied once, before either nested model, as a
diagnostic control.

```{r lingoes-control}
lingoes_constant <- function(dm) {
  d2 <- as.matrix(dm)^2
  n <- nrow(d2)
  J <- diag(n) - matrix(1 / n, nrow = n, ncol = n)
  G <- -0.5 * J %*% d2 %*% J
  values <- eigen((G + t(G)) / 2, symmetric = TRUE, only.values = TRUE)$values
  max(-min(values), 0)
}
bray_lingoes_constant <- lingoes_constant(D_bray)
D_bray_lingoes <- sqrt(D_bray^2 + 2 * bray_lingoes_constant)
attr(D_bray_lingoes, "method") <- "Lingoes-corrected Bray-Curtis"
lingoes_result <- evaluate_distance("Lingoes-corrected Bray-Curtis",
  D_bray_lingoes, simulated_metadata)
knitr::kable(rbind(
  distance_sensitivity[distance_sensitivity$distance_definition == "Bray-Curtis", ],
  lingoes_result), digits = 4)
```

After this one-time correction, vegan and WdStar one-pass agree because no
substantive negative axes need removal. WdStar two-pass still differs. The projection
problem therefore exists independently of the negative-eigenvalue problem.

# Real-data illustrations

The simulated data remain the primary example. This section shows how the same
diagnostics apply to variables in the `enterotype` dataset.

## Enterotype adjusted for sequencing technology

```{r enterotype-data}
data("enterotype")
ent <- subset_samples(enterotype, !is.na(Enterotype) & !is.na(SeqTech))
metadata <- data.frame(sample_data(ent))
D_enterotype <- distance(ent, method = "bray")
enterotype_table <- table(SeqTech = metadata$SeqTech,
  Enterotype = metadata$Enterotype)
enterotype_chisq <- suppressWarnings(chisq.test(enterotype_table, correct = FALSE))
enterotype_cramers_v <- sqrt(as_number(enterotype_chisq$statistic) /
  (sum(enterotype_table) *
     min(nrow(enterotype_table) - 1, ncol(enterotype_table) - 1)))
knitr::kable(enterotype_table, caption = "SeqTech by Enterotype")
knitr::kable(data.frame(
  chi_squared = as_number(enterotype_chisq$statistic),
  degrees_of_freedom = as_number(enterotype_chisq$parameter),
  p_value = format.pval(enterotype_chisq$p.value, digits = 3, eps = 0.001),
  Cramers_V = enterotype_cramers_v), digits = 4)
```

This tests predictor association, not Bray--Curtis or microbiome model fit. It shows
why term order and shared predictor information are relevant.

```{r enterotype-comparison}
D_ent_F <- a.dist(D_enterotype, ~ Enterotype, metadata)
D_ent_Z <- a.dist(D_enterotype, ~ SeqTech, metadata)
D_ent_full <- a.dist(D_enterotype, ~ SeqTech + Enterotype, metadata)
D_ent_two <- a.dist(D_ent_Z, ~ Enterotype, metadata)
ent_adonis <- vegan::adonis2(D_enterotype ~ SeqTech + Enterotype,
  data = metadata, by = "terms", permutations = 999)
ent_f_only <- vegan::adonis2(D_enterotype ~ Enterotype,
  data = metadata, by = "terms", permutations = 999)

ent_r2_f <- as_number(dist.goodness.of.fit(D_enterotype, D_ent_F))
ent_r2_z <- as_number(dist.goodness.of.fit(D_enterotype, D_ent_Z))
ent_r2_one <- as_number(dist.goodness.of.fit(D_enterotype, D_ent_full))
ent_r2_two <- as_number(dist.goodness.of.fit(D_enterotype, D_ent_two))
ent_vegan_f <- as_number(ent_f_only["Enterotype", "R2"])
ent_vegan_z <- as_number(ent_adonis["SeqTech", "R2"])
ent_vegan_semi <- as_number(ent_adonis["Enterotype", "R2"])
ent_vegan_resid <- as_number(ent_adonis["Residual", "R2"])
ent_vegan_full <- 1 - ent_vegan_resid

enterotype_comparison <- data.frame(
  quantity = c("Enterotype-only R2", "SeqTech adjustment-only R2",
    "Full/cumulative R2: SeqTech + Enterotype",
    "Semi-partial R2: Enterotype after SeqTech",
    "Partial R2: Enterotype after SeqTech"),
  vegan_shared_geometry = c(ent_vegan_f, ent_vegan_z, ent_vegan_full,
    ent_vegan_semi, ent_vegan_semi / (1 - ent_vegan_z)),
  WdStar_one_pass = c(ent_r2_f, ent_r2_z, ent_r2_one,
    ent_r2_one - ent_r2_z,
    (ent_r2_one - ent_r2_z) / (1 - ent_r2_z)),
  WdStar_two_pass = c(NA, ent_r2_z, ent_r2_two,
    ent_r2_two - ent_r2_z,
    (ent_r2_two - ent_r2_z) / (1 - ent_r2_z)))
knitr::kable(enterotype_comparison, digits = 4,
  caption = "Same samples and Bray-Curtis distance: F = Enterotype; Z = SeqTech")
```

## Age as another adjustment variable

`Age` is continuous, so a chi-square test and Cramér's $V$ are not appropriate. Group
summaries can describe its relationship with `Enterotype`; the distance model uses
the same shared-geometry nested comparison. This shows that the issue is not limited
to two categorical variables.

```{r age-illustration}
ent_age <- subset_samples(enterotype, !is.na(Enterotype) & !is.na(Age))
age_metadata <- data.frame(sample_data(ent_age))
D_age <- distance(ent_age, method = "bray")
age_summary <- aggregate(Age ~ Enterotype, data = age_metadata,
  FUN = function(x) c(n = length(x), mean = mean(x), sd = sd(x)))
age_summary <- data.frame(Enterotype = age_summary$Enterotype,
  n = age_summary$Age[, "n"], mean_age = age_summary$Age[, "mean"],
  sd_age = age_summary$Age[, "sd"])
age_adonis <- vegan::adonis2(D_age ~ Age + Enterotype,
  data = age_metadata, by = "terms", permutations = 999)
knitr::kable(age_summary, digits = 2)
knitr::kable(age_adonis, digits = 4)
```

With `Age` first, the `Enterotype` row is its semi-partial contribution after linear
adjustment for age. A second call to `a.dist()` is still unnecessary.

# $W_d^*$ is a separate result

$W_d^*$ and `dist.goodness.of.fit()` are different statistics and should not be
presented as interchangeable estimates.

```{r wds-examples}
WdS.test(dm = D_enterotype, f = metadata$Enterotype)
dist.goodness.of.fit(D_enterotype, D_ent_F)
WdS.test(dm = D_enterotype, f = metadata$Enterotype,
  formula = ~ SeqTech, formula_data = metadata)
```

The goodness-of-fit value printed for the adjustment step describes the change from
the original distance to the `SeqTech`-adjusted distance. It is $R_Z^2$, not the
conditional contribution of `Enterotype` after `SeqTech`.

# Conclusions and implementation direction

1. Calculate $TSS$, $RSS_Z$, and $RSS_{ZF}$ from one shared Gower matrix.
2. Report semi-partial $F$ after $Z$ as $(RSS_Z-RSS_{ZF})/TSS$, and partial $F$ after
   $Z$ as $(RSS_Z-RSS_{ZF})/RSS_Z$.
3. Do not use `a.dist(a.dist(D, ~ Z), ~ F)` as the standard sequential fit. Its final
   residual generally differs from the joint residual even for a Euclidean distance.
4. Label the exact distance: quantitative versus binary Jaccard, and raw
   Jensen--Shannon divergence versus its square-root distance.
5. Diagnose negative eigenvalues. If Euclidification is desired, apply one declared
   correction to the original distance before fitting every nested model.
6. Changing distance changes the scientific target and numerical behavior.
   Bray--Curtis emphasizes abundance differences, binary Jaccard presence/absence,
   and Jensen--Shannon relative-abundance distributions.
7. Residual distances may be useful for visualization, but model $R^2$ should use a
   shared signed geometry rather than separately reconstructed residual distances.

# Session information

```{r session-information}
sessionInfo()
```
